---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# PerturbPlan

<!-- badges: start -->

[![R-CMD-check](https://github.com/Katsevich-Lab/perturbplan/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Katsevich-Lab/perturbplan/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/Katsevich-Lab/perturbplan/branch/main/graph/badge.svg)](https://app.codecov.io/gh/Katsevich-Lab/perturbplan?branch=main)

<!-- badges: end -->

**PerturbPlan** is an R package for power analysis in perturb-seq and TAP-seq experiments. It complements the [PerturbPlan web app](https://katsevich-lab-perturbplan.share.connect.posit.cloud/), which provides an interactive interface for experimental design. 

# Use Cases

The PerturbPlan package supports three use cases:

1. **[Prepare Reference Data for Web App](articles/preprocess-reference.html)** - Shiny app users wishing to provide custom reference data to the app should use the package to prepare their data before upload.
2. **[Advanced Prospective Power Analysis](articles/prospective-power.html)** - While the app helps users address 11 of the most commonly encountered experimental design questions, the package supports more sophisticated analyses.
3. **[Retrospective Power Analysis](articles/posthoc.html)** - In addition to experimental design---based on prospective power analysis---the package also supports retrospective  power analysis.

## Installation

You can install PerturbPlan from GitHub:

```{r eval=FALSE}
# install.packages("devtools")
devtools::install_github("Katsevich-Lab/perturbplan")
```
