---
title: "Posthoc Power Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Posthoc Power Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "vignettes/figures/posthoc-",
  out.width = "100%"
)
```

In the example below, we will demonstrate how to use `compute_power_posthoc()` to compute power after an experiment is conducted (as opposed to prospectively).

```{r}
library(perturbplan)
```


```{r, echo = FALSE}
set.seed(1)

num_enhancers <- 3
grnas_per_enhancer <- 2
num_genes <- 4

enhancers <- c(paste0("enh", 1:num_enhancers), "nt")
grnas <- paste0("grna", 1:grnas_per_enhancer)
genes <- paste0("gene", 1:num_genes)
cells_per_grna <- expand.grid(enhancer = enhancers, grna = grnas) |>
  dplyr::mutate(grna_id = paste0(enhancer, "_", grna)) |>
  dplyr::rename(grna_target = enhancer) |> 
  dplyr::select(grna_id, grna_target) |>
  dplyr::mutate(num_cells = rpois(dplyr::n(), lambda = 100),
                grna_target = dplyr::if_else(grna_target == "nt", "non-targeting", grna_target))
  
discovery_pairs <- within(expand.grid(
  grna_target = c("enh1","enh2","enh3"),
  response_id = paste0("gene", 1:4),
  KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE
), { grna_target <- as.character(grna_target); response_id <- as.character(response_id) })

baseline_expression_stats <- data.frame(
  response_id = genes,
  expression_mean = exp(rnorm(num_genes, mean = 1, sd = 1)),
  expression_size = exp(rnorm(num_genes, mean = 1, sd = 1))
)
```

## Posthoc power inputs

### 1. Dataset summaries

Let us consider a toy perturb-seq dataset with 3 enhancers, each targeted by 2 gRNAs, 2 non-targeting gRNAs, and 4 genes. Suppose there are 1000 total cells in the experiment:
```{r}
num_total_cells <- 1000
```
Based on our data, suppose we have information on the number of cells that received each gRNA:
```{r}
cells_per_grna
```
Furthermore, we have computed the mean and size parameters for the baseline expression of each gene:
```{r}
baseline_expression_stats
```

### 2. Analysis choices

Suppose we tested the effects of all elements on all genes. We can encode this in a `discovery_pairs` data frame, the same as in `sceptre`:
```{r}
discovery_pairs
```
We have analyzed the data using the complement control group, left-sided tests, and default values for the pairwise QC parameters:
```{r}
control_group <- "complement"
side <- "both"
n_nonzero_trt_thresh <- 7
n_nonzero_cntrl_thresh <- 7
```

### 3. Power analysis parameters

We want to compute the power for enhancer-gene pair assuming that gRNAs targeting that enhancer have fold changes on the gene with mean and standard deviations of 0.85 and 0.13, respectively:
```{r}
fold_change_mean <- 0.85
fold_change_sd <- 0.13
```
We want to deem perturbation-gene pairs as significant in our power analysis if they pass a threshold of 0.005:
```{r}
cutoff <- 0.005
```
This number may be obtained as the $p$-value cutoff from the original analysis. 

## Running PerturbPlan posthoc power analysis

We can now compute the power for each enhancer-gene pair using the `compute_power_posthoc()` function:
```{r}
power_results <- compute_power_posthoc(
  num_total_cells = num_total_cells,
  cells_per_grna = cells_per_grna,
  baseline_expression_stats = baseline_expression_stats,
  discovery_pairs = discovery_pairs,
  control_group = control_group,
  side = side,
  n_nonzero_trt_thresh = n_nonzero_trt_thresh,
  n_nonzero_cntrl_thresh = n_nonzero_cntrl_thresh,
  fold_change_mean = fold_change_mean,
  fold_change_sd = fold_change_sd,
  cutoff = cutoff
)
```

## PerturbPlan posthoc power outputs

The result is a list with fields `expected_num_discoveries` and `individual_power`:
```{r}
names(power_results)
```
The field `expected_num_discoveries` contains the expected number of discoveries across all enhancer-gene pairs:
```{r}
power_results$expected_num_discoveries
```
The field `individual_power` is a data frame containing the power for each enhancer-gene pair:
```{r}
power_results$individual_power
```

