---
title: "Preprocess Custom Reference Expression Data for Web App"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preprocess Custom Reference Expression Data for Web App}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows how to create custom reference data for PerturbPlan. The process has three steps:

1. Preprocess your data
2. Extract summary statistics from preprocessed data
3. Save the result

```{r setup}
library(perturbplan)
```

## Step 1: Preprocess Your Data

The goal is to preprocess your data into the format expected by `reference_data_processing()`. There are two cases.

### Case 1: You Have Cell Ranger Output

If your data is organized as Cell Ranger run directories, use `reference_data_preprocessing_10x()` to extract the inputs needed by `reference_data_processing()`.

#### Required Layout

The convenience helper expects Cell Ranger **count** output arranged under a top-level directory:

```text
path_to_top_level_output/
├── SRR_run_1/
│   └── outs/
│       ├── filtered_feature_bc_matrix/
│       │   ├── barcodes.tsv.gz
│       │   ├── features.tsv.gz
│       │   └── matrix.mtx.gz
│       ├── molecule_info.h5
│       ├── filtered_feature_bc_matrix.h5
│       └── metrics_summary.csv
├── SRR_run_2/
│   └── ...
└── SRR_run_3/
    └── ...
```

`reference_data_preprocessing_10x()` is currently designed for Cell Ranger `count`, not Cell Ranger `multi`.

#### Run `reference_data_preprocessing_10x()`

We illustrate this function with a tiny example directory, `cellranger_tiny`.

```{r preprocess-10x}
extdata_path <- system.file("extdata", package = "perturbplan")

preprocessed_data <- reference_data_preprocessing_10x(
  path_to_top_level_output = extdata_path,
  path_to_run_level_output = "cellranger_tiny",
  h5_rough = TRUE
)

names(preprocessed_data)
str(preprocessed_data, max.level = 1)
```

After this step, `preprocessed_data` contains the three objects consumed by `reference_data_processing()`:

- `response_matrix`
- `read_umi_table`
- `mapping_efficiency`

### Case 2: You Do Not Have Cell Ranger Output

If your data does not come from Cell Ranger, you can still use PerturbPlan as long as you can provide compatible inputs to `reference_data_processing()`.

At minimum:

- `response_matrix` should be a gene-by-cell matrix with genes in rows and cells in columns
- `read_umi_table` should contain UMI-level molecule information in the format expected by the library-fitting step
- `mapping_efficiency` is optional but recommended when available

In this case, manipulate your data yourself into the required format, checking `?reference_data_processing` for the expected input-format details. Once those objects are ready, proceed to Step 2.

## Step 2: Extract Summary Statistics From Preprocessed Data

The examples below continue using the `preprocessed_data` object created in Step 1, Case 1. If you prepared compatible inputs yourself in Step 1, Case 2, the same call pattern applies to your own `response_matrix`, `read_umi_table`, and optional `mapping_efficiency`.

There are two common cases. In both cases, the summary statistics are extracted from the preprocessed data via `reference_data_processing()`.

### Case 1: Perturb-seq

Use `gene_list = NULL` to keep all genes. If desired, select a TPM threshold for genes to include in processing. Here, we illustrate using a threshold of `0.1`.

```{r process-perturbseq}
set.seed(123)

pilot_data_perturbseq <- reference_data_processing(
  response_matrix = preprocessed_data$response_matrix,
  read_umi_table = preprocessed_data$read_umi_table,
  mapping_efficiency = preprocessed_data$mapping_efficiency,
  gene_list = NULL,
  TPM_thres = 0.1
)

names(pilot_data_perturbseq)
```

### Case 2: TAP-seq

If your reference expression data contain more genes than your planned TAP-seq experiment, use `gene_list` to restrict the analysis to a targeted panel. In that setting, `TPM_thres = 0` is often a reasonable choice because the target panel is already preselected.

```{r process-targeted}
gene_list <- c("ENSG00000241860", "ENSG00000238009", "ENSG00000239945")

pilot_data_targeted <- reference_data_processing(
  response_matrix = preprocessed_data$response_matrix,
  read_umi_table = preprocessed_data$read_umi_table,
  mapping_efficiency = preprocessed_data$mapping_efficiency,
  gene_list = gene_list,
  TPM_thres = 0
)

names(pilot_data_targeted)
```

## Step 3: Save the Result

The output of `reference_data_processing()` is your custom reference dataset. Save that object as an RDS file:

```{r save-rds, eval=FALSE}
saveRDS(pilot_data_perturbseq, "my_perturbplan_reference.rds")
```

This RDS file is the file you will upload to the [PerturbPlan web app](perturbplan.com) as your custom reference dataset.

## See Also

- `?reference_data_preprocessing_10x`
- `?reference_data_processing`
